% =====================================================================
%  Reference-list entries (LaTeX \bibitem form) for the foundational
%  references introduced in the Theory chapter (theory_chapter.R).
%
%  USAGE
%  -----
%  If your thesis uses a manual reference list, paste the \bibitem
%  lines below into your existing thebibliography environment.
%  If you have NO reference list yet, keep the surrounding
%  \begin{thebibliography}...\end{thebibliography} as well.
%
%  The optional argument in \bibitem[...]{key} is the author-year label
%  shown in the text; it is used by natbib-style author-year citations.
%  If you use plain numeric citations, you may delete the [...] part.
%  The \bibitem keys match \cite keys in theory_chapter.R exactly.
% =====================================================================

\begin{thebibliography}{99}

\bibitem[Brier(1950)]{brier1950verification}
Brier, G.~W. (1950).
\newblock Verification of forecasts expressed in terms of probability.
\newblock \emph{Monthly Weather Review}, 78(1), 1--3.

\bibitem[Cohen(1988)]{cohen1988statistical}
Cohen, J. (1988).
\newblock \emph{Statistical Power Analysis for the Behavioral Sciences} (2nd ed.).
\newblock Hillsdale, NJ: Lawrence Erlbaum Associates.

\bibitem[Efron \& Tibshirani(1993)]{efron1993introduction}
Efron, B., \& Tibshirani, R.~J. (1993).
\newblock \emph{An Introduction to the Bootstrap}.
\newblock New York: Chapman and Hall.

\bibitem[Fisher(1925)]{fisher1925statistical}
Fisher, R.~A. (1925).
\newblock \emph{Statistical Methods for Research Workers}.
\newblock Edinburgh: Oliver and Boyd.

\bibitem[Good(2005)]{good2005permutation}
Good, P.~I. (2005).
\newblock \emph{Permutation, Parametric, and Bootstrap Tests of Hypotheses} (3rd ed.).
\newblock New York: Springer.

\bibitem[Hanley \& McNeil(1982)]{hanley1982meaning}
Hanley, J.~A., \& McNeil, B.~J. (1982).
\newblock The meaning and use of the area under a receiver operating characteristic (ROC) curve.
\newblock \emph{Radiology}, 143(1), 29--36.

\bibitem[Hodges \& Reich(2010)]{hodges2010adding}
Hodges, J.~S., \& Reich, B.~J. (2010).
\newblock Adding spatially-correlated errors can mess up the fixed effect you love.
\newblock \emph{The American Statistician}, 64(4), 325--334.

\bibitem[K\"{u}nsch(1989)]{kunsch1989jackknife}
K\"{u}nsch, H.~R. (1989).
\newblock The jackknife and the bootstrap for general stationary observations.
\newblock \emph{The Annals of Statistics}, 17(3), 1217--1241.

\bibitem[Murphy(1973)]{murphy1973new}
Murphy, A.~H. (1973).
\newblock A new vector partition of the probability score.
\newblock \emph{Journal of Applied Meteorology}, 12(4), 595--600.

\bibitem[Reich et~al.(2006)]{reich2006effects}
Reich, B.~J., Hodges, J.~S., \& Zadnik, V. (2006).
\newblock Effects of residual smoothing on the posterior of the fixed effects in disease-mapping models.
\newblock \emph{Biometrics}, 62(4), 1197--1206.

\bibitem[Spearman(1904)]{spearman1904proof}
Spearman, C. (1904).
\newblock The proof and measurement of association between two things.
\newblock \emph{The American Journal of Psychology}, 15(1), 72--101.

\end{thebibliography}
